Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-03-23 content.tex
3123 words

We use the tree as shown in figure \ref{fig:dd_tree}.
The with six taxa, to keep the figure more readable,
with an arbitrary crown age of ten time units.

@@

The first step in \verb;pirouette; is to 
simulate a DNA alignment from the 
given phylogeny, as described in Subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 
mononucleotides each, while the per-nucleotide mutation rate is 
0.1 mutations per unit time.
Together with the crown age of 10 time units, this results
in an expected per-locus mutation probability of 1.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) NSM
and a strict clock model as these are the simplest.
A JC model assumes that mutation rates between nucleotides are equal and 
constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

In the second step we state our experiment.
We define an experiment $\mathit{X}$ 
as a combination of an inference model 
$\mathit{I}$ and conditions $\mathit{C}$.
As a generative model, we assume a
site model with a JC69 NSM,
a strict clock model and a Yule tree prior.
For our candidate models,
we select all 16 experiments
that are a combination of the 4 NSMs,
2 clock models and 2 BD tree priors.
From this set, we exclude the generative model.

We use the nLTT statistic to
measure the difference between phylogenies,
with a burn-in of 10\%.

@@

While the error distributions using the best or generative model 
as inference model are very similar, 
the error distributions of the true tree are substantially 
larger than those of the twin tree. 
This is the error made by the mismatch
between the generating species tree model and the tree prior used in inference.

@@

Figure~\ref{fig:example_3} illustrates the primary result of our pipeline: 
it shows the error distributions for the true tree and the twin tree 
when either the generating model or the best candidate model is used in inference. 
The clear difference between the error distributions 
for the true tree and the twin tree suggests 
that the choice of tree prior does matter.

@@

\verb;pirouette; is free, libre and open source software available at 
\url{http://github.com/richelbilderbeek/pirouette},
licensed under the GNU General Public License version 3.
\verb;pirouette; depends on multiple packages, which are:
\verb;ape; (\cite{ape}),
\verb;babette; (\cite{bilderbeek2018babette}),
\verb;becosys; (\cite{becosys}),
\verb;DDD; (\cite{DDD}),
\verb;devtools; (\cite{devtools}),
\verb;dplyr; (\cite{dplyr}),
\verb;geiger; (\cite{geiger}),
\verb;ggplot2; (\cite{ggplot2}),
\verb;knitr; (\cite{knitr}),
\verb;lintr; (\cite{lintr}),
\verb;magrittr; (\cite{magrittr}),
\verb;mcbette; (\cite{mcbette}),
\verb;nLTT; (\cite{nLTT}),
\verb;PBD; (\cite{PBD}),
\verb;phangorn; (\cite{phangorn}),
\verb;phytools; (\cite{phytools}),
\verb;plyr; (\cite{plyr}),
\verb;rappdirs; (\cite{rappdirs}),
\verb;rmarkdown; (\cite{rmarkdown}),
\verb;Rmpfr; (\cite{Rmpfr}),
\verb;stringr; (\cite{stringr}),
\verb;TESS; (\cite{TESS}),
\verb;testit; (\cite{testit}), 
\verb;testthat; (\cite{testthat}) and
\verb;tidyr; (\cite{tidyr}).

>>>>

We use the tree as shown in figure \ref{fig:dd_tree},
which has an arbitrary crown age of ten time units
and six taxa, to keep the figure more readable.

@@

Using the default \verb;pirouette; settings,
the error distribution shown in figure \label{fig:example_30}
is produced.

@@

Focussing on the upper panel of figure \ref{fig:example_30},
we can see that the error distributions of the generative model
differ strongly between the true and twin tree. 
This difference shows the extent of the mismatch between
the actual tree model (which is DD) and the (BD) tree prior used.
Because these distributions are distinctively different,
the inference error we make when using an 
incorrect (that is, BD) tree prior on a DD tree
is profound.

Comparing the upper and lower panel of figure \ref{fig:example_30}, 
we can see that the best
candidate model is only slightly better at inferring the true tree,
compared to the assumed generative model, confirming that the 
picked generative model was a good choice.

@@

Figure~\ref{fig:example_30} illustrates the primary result of our pipeline: 
it shows the error distributions for the true tree and the twin tree 
when either the generating model or the best candidate model is used in inference. 
The clear difference between the error distributions 
for the true tree and the twin tree suggests 
that the choice of tree prior does matter.

@@

\verb;pirouette; is free, libre and open source software available at 
\url{http://github.com/richelbilderbeek/pirouette},
licensed under the GNU General Public License version 3.
\verb;pirouette; depends on multiple packages, which are:
\verb;ape; (\cite{ape}),
\verb;assertive; (\cite{assertive}),
\verb;babette; (\cite{bilderbeek2018babette}),
\verb;DDD; (\cite{DDD}),
\verb;devtools; (\cite{devtools}),
\verb;dplyr; (\cite{dplyr}),
\verb;ggplot2; (\cite{ggplot2}),
\verb;knitr; (\cite{knitr}),
\verb;lintr; (\cite{lintr}),
\verb;magrittr; (\cite{magrittr}),
\verb;mcbette; (\cite{mcbette}),
\verb;nLTT; (\cite{nLTT}),
\verb;phangorn; (\cite{phangorn}),
\verb;phytools; (\cite{phytools}),
\verb;plyr; (\cite{plyr}),
\verb;rappdirs; (\cite{rappdirs}),
\verb;rmarkdown; (\cite{rmarkdown}),
\verb;Rmpfr; (\cite{Rmpfr}),
\verb;stringr; (\cite{stringr}),
\verb;TESS; (\cite{TESS}),
\verb;testit; (\cite{testit}), 
\verb;testthat; (\cite{testthat}) and
\verb;tidyr; (\cite{tidyr}).
